### 2014-11-08 abstract.tex
More updates.

Declarative programming has been hailed as a promising approach to parallel
programming since it hides the implementation details of parallelism away from the
programmer. However, some control is lost when code is
automatically parallelized, leaving the programmer with very little
opportunities for fine tuning programs.
We extended Linear Meld, a concurrent forward-chaining linear logic programming
language, with coordination facts that allow the programmer to change
computation scheduling and data layout. Our approach allows the programmer to
write declarative parallel programs and then optionally use coordination facts
to fine tune programs, while keeping the program fully declarative and amenable
to correctness proofs. We show several application examples, experimental
results and some advantages of our novel approach.

>>>>

Declarative programming has been hailed as a promising approach to parallel
programming since it hides the implementation details of parallelism away from the
programmer. However, its advantage has also been its downfall
as it leaves the programmer with no straightforward way
to optimize programs for performance.
In this paper, we introduce Coordinated Linear Meld, a concurrent forward-chaining linear logic programming
language, with a declarative way to coordinate the execution of the program
allowing the programmer to change
computation scheduling and data layout. Our approach allows the programmer to
write declarative parallel programs and then optionally use coordination 
to fine tune programs, while keeping the program fully declarative and amenable
to correctness proofs.

### 2014-11-08 application.tex
New update.
More updates.

To better understand how coordination facts are used, we next present some programs that
take advantage of them.

@@

\subsection{MiniMax}

\input{minimax}

\subsection{Heat Transfer}

\input{heat-transfer}

\subsection{Splash Belief Propagation}

\input{splash-bp}

\subsection{N Queens}

>>>>

To better understand how coordination facts are used, we next present some programs that
take advantage of them. In our experimental setup, we used a machine with
four AMD Six-Core Opteron TM 8425 HE (2100 MHz) chips (24 cores) and 64 GB of
DDR-2 667MHz (16x4GB) RAM, running GNU/Linux (kernel 3.15.10-201 64 bits).
We compiled our virtual machine using GCC 4.8.3 (g++) with the flags
\texttt{-O3 -std=c++11 -fno-rtti -march=x86-64}~\footnote{Implementation
   available in \url{http://github.com/.../meld}}.

@@

\subsection{MiniMax}
\input{minimax}

\subsection{Heat Transfer}
\input{heat-transfer}

\subsection{Splash Belief Propagation}
\input{splash-bp}

\subsection{N Queens}
\input{nqueens}

### 2014-11-08 coordination.tex
More corrections.
More updates.

Since LM uses linear logic and supports deletion of facts, the order in which
nodes are scheduled can impact the performance and even the results of the
program.

The SSSP program present in Fig.~\ref{code:shortest_path_program} is concise and
declarative but its performance may depend in the order in which nodes are
executed. If nodes with greater distances are prioritized over other nodes, the
program will generate more \texttt{relax} facts since it will take longer to
reach the shortest distances. From Fig.~\ref{fig:shortest_path_program}, it is
clear that the best computational ordering is the following: \texttt{@1},
\texttt{@3}, \texttt{@2} and then \texttt{@4}, where only 4 \texttt{relax}
facts are generated. If we had decided to process nodes in order
\texttt{@1}, \texttt{@2}, \texttt{@4}, \texttt{@3}, \texttt{@4},
\texttt{@2}, then 6 \texttt{relax} facts would have been generated.
Therefore, the optimal solution for SSSP is to schedule the node with the
shortest distance, which is essentially the Dijkstra shortest path
algorithm~\cite{Dijkstra}. Note how it is possible to change the nature of
the algorithm by simply changing the order of node computation, but still
retain the declarative nature of the program.

We introduce the concept of \emph{coordination facts} that allow the programmer
to change how the run time schedules nodes and partitions the nodes among
threads of execution. Coordination facts can be used in the body or head of the
rules to allow the programmer to reason and change how execution is to be done.
In this fashion, we keep the language declarative because we reason logically
about the state of execution, without the need to introduce extra-logical
operators into the language that would introduce problems when attempting to prove
properties about programs.

There are two classes of coordination facts. The first class of coordination facts are
called \emph{sensing facts} and are used to sense information about the
underlying runtime system, including node placement and node scheduling.
The second kind of coordination facts are \emph{action facts} that are deleted
to apply a coordination operation in the runtime system.

@@

We can use action facts to change the order in which nodes are evaluated by adding
\emph{priorities}. Node priority works at the thread level
so that each thread can make a local decision about which node to execute next.
Note that, by default, nodes are picked using a FIFO approach, because that
tends to work better for most programs since older nodes tend to have more facts
to be used.

@@

\begin{itemize}
   \item \texttt{type linear set-priority(node, float)}: This sets the
   temporary priority of a node. If the node already has a priority, we only
   change the priority if the new one is higher. The programmer can
   decide if priorities are to be ordered in ascending or descending order.
   \item \texttt{type linear action add-priority(node, float)}: Increases,
   temporarily, the priority of the node.
   \item\texttt{type linear schedule-next(node)}: The work will fetch the
   highest priority node's priority $P$ from its set of nodes and set the
   action's argument node's priority as $P + 1.0$. If using the priorities in
   ascending order, we pick the lowest priority and subtract $1.0$.
   \item\texttt{type linear set-default-priority(node, float)}: Sets the default
   priority of the node.
   \item \texttt{type linear action stop-program(node)}: Immediately stops the
   execution of the whole program.
\end{itemize}

LM provides the sensing fact \texttt{priority(A, B, P)} in order to sense the
priority \texttt{P} of node \texttt{B} from node \texttt{A}.
Sensing facts are only used in the body of the rules in order
to fetch information from the runtime system.
Note that when sensing facts are deleted, they are re-derived automatically.
Logically, \texttt{set-priority} and \texttt{set-default-priority} update the
value of \texttt{priority} facts, but this done automatically by the runtime
system.

@@

\begin{itemize}
   \item \texttt{type linear set-cpu(node, int)}: Moves the node to a
   thread of execution.
   \item \texttt{type linear set-affinity(node, node)}: Places the first node in
   the same thread as the second node.
   \item \texttt{type linear set-moving(node)}: Allows the node to move freely
   between threads.
   \item \texttt{type linear set-static(node)}: Forces the node to stay in the
   same thread indefinitely.
\end{itemize}

@@

\begin{itemize}
   \item \texttt{type linear cpu-id(node, node, int)}: The third argument
   indicates the thread where the node of the second argument is currently running.
   \item \texttt{type linear moving(node, node)}: Fact available if the node in the
   second argument is allowed to move between threads.
   \item \texttt{type linear static(node, node)}: Fact available if the node in
   the second argument is not allowed to move between threads.
\end{itemize}

>>>>

The SSSP program present in Fig.~\ref{code:shortest_path_program} is concise and
declarative but its performance depends on the order in which nodes are
executed. If nodes with greater distances are prioritized over other nodes, the
program will generate more \texttt{relax} facts since it will take longer to
reach the shortest distances. From Fig.~\ref{fig:shortest_path_program}, it is
clear that the best scheduling is the following: \texttt{@1},
\texttt{@3}, \texttt{@2} and then \texttt{@4}, where only 4 \texttt{relax}
facts are generated. If we had decided to process nodes in order
\texttt{@1}, \texttt{@2}, \texttt{@4}, \texttt{@3}, \texttt{@4},
\texttt{@2}, then 6 \texttt{relax} facts would have been generated.
The optimal solution for SSSP is to schedule the node with the
shortest distance, which is essentially the Dijkstra shortest path
algorithm~\cite{Dijkstra}. Note how it is possible to change the nature of
the algorithm by simply changing the order of node computation, but still
retain the declarative nature of the program.

\emph{Coordination facts} allow the programmer
to change how the run time schedules nodes and how it partitions the nodes among
threads of execution. Coordination facts can be used in either the body of the
rule, the head of the rule or both.
This allows scheduling and partition decisions to be made based on the state of
the program and on the state of the underlying machine.
In this fashion, we keep the language declarative because we reason logically
about the state of execution, without the need to introduce extra-logical
operators into the language that would introduce significant issues when proving
properties about programs.

There are two kinds of coordination facts: \emph{sensing} and \emph{action}
facts. Sensing facts are used to sense information about the
underlying runtime system, including node placement and node scheduling.
Action facts are used to apply a coordination operations on the runtime system.

@@

In order to allow different scheduling strategies, we introduce the concept of
\emph{node priority} by assigning a priority value to every node in the program
and by introducing coordination facts that manipulate such priority values.
By default, nodes do not have any priority and can be picked in any order. In
our implementation, we use a FIFO approach because this tends to work better for
most programs since older nodes tend to a higher number of unexamined facts.

@@

\newcommand{\code}[1]{\texttt{\small{#1}}}

\begin{itemize}
   \item \code{set-priority(node A, float F)}: This sets the
   temporary priority of \texttt{A} to \texttt{F}. If \texttt{A} has a priority
   \texttt{F'}, we only change the priority if \texttt{F} is higher. The programmer
   can decide if priorities are to be ordered in ascending or descending order.
   \item \code{add-priority(node A, float F)}: Increases,
   temporarily, the priority of node \texttt{A} by \texttt{F}.
   \item \code{schedule-next(node A)}: Changes the temporary priority of node
   \texttt{A} to be 1 unit higher than the current highest priority.
   \item \code{set-default-priority(node, float)}: Sets the default
   priority of the node.
   \item \code{stop-program(node A)}: Immediately stops the
   execution of the whole program.
\end{itemize}

LM provides the sensing fact \code{priority(node A, node B, float P)} in order to sense the
priority \texttt{P} of node \texttt{B} (at node \texttt{A}).
Sensing facts are only used in the body of the rules in order
to fetch information from the runtime system.
Note that when sensing facts are used to prove new facts, they are re-derived automatically.
~\footnote{The observant reader will know that all the coordination facts are
   linear and XXX
Logically, \texttt{set-priority} and \texttt{set-default-priority} update the
value of \texttt{priority} facts, but this done automatically by the runtime
system.}

@@

\begin{itemize}
   \item \code{set-cpu(node A, int C)}: Moves node \texttt{A} to thread
   \texttt{C}.
   \item \code{set-affinity(node A, node B)}: Places node \texttt{B} in
   the thread of node \texttt{B}.
   \item \code{set-moving(node A)}: Allows node \texttt{A} to move freely
   between threads.
   \item \code{set-static(node A)}: Forces node \texttt{A} to stay in the
   same thread indefinitely.
\end{itemize}

@@

\begin{itemize}
   \item \code{cpu-id(node A, node B, int C)}: Store at node \texttt{A} which
   thread, \texttt{C}, \texttt{B} is actually running on.
   \item \code{moving(node A, node B)}: Fact available at node \texttt{A} if \texttt{B} is allowed
   to move between threads.
   \item \code{static(node A, node B)}: Fact available at node \texttt{A} if \texttt{B} is not
   allowed to move between threads.
\end{itemize}

### 2014-11-08 heat-transfer.tex
Small changes.
More updates.

In the Heat Transfer algorithm~(HT), we have a graph where heat values are
exchanged between nodes. The goal of the program is to exchange heat to a point
where new heat value $\delta = |H_i -
H_{i-1}| \le \epsilon$ for all nodes of the graph. The algorithm works
asynchronously since heat values can be updated by using new partial information
coming from neighbor nodes. This increases parallelism since nodes do not need
to synchronize if we had to compute heat values using iterations.

Fig.~\ref{code:ht} shows the HT logical rules that send new heat values to
neighbor nodes. We added the first rule to increase the priority of the neighbor
nodes if the current node has a large $\delta$. The idea is to prioritize the
computation of heat values of nodes (using \texttt{update}) that have a neighbor
node that changed significantly. Multiple \texttt{add-priority} facts will
increase the priority of a node so that nodes with multiple large deltas will
have more priority. Note that the LM compiler will merge the three rules in
Fig.~\ref{code:ht} into a single one that handles the 3 branches.

\begin{figure}[h!]
\scriptsize\begin{Verbatim}[numbers=left,commandchars=\\\[\]]
new-heat(A, New, Old),
Delta = fabs(New - Old),
Delta > epsilon * 2.0
   -o {B, W | !edge(A, B, W) |
         new-neighbor-heat(B, A, New),
         update(B), \underline[add-priority(B, Delta)]}.

new-heat(A, New, Old),
Delta = fabs(New - Old)
Delta < epsilon * 2.0,
Delta > epsilon
   -o {B, W | !edge(A, B, W) |
         new-neighbor-heat(B, A, New), update(B)}.

new-heat(A, New, Old)
Delta = fabs(New - Old)
Delta < epsilon
   -o {B, W | !edge(A, B, W) |
         new-neighbor-heat(B, A, New)}.
\end{Verbatim}
  \caption{Coordination code for the Heat Transfer program. We added one rule
     that covers cases where $\Delta$ is at least $2 \epsilon$ in order to
     increase the priority of neighbors nodes. The code logic remains exactly
     the same as before, however bigger changes in heat values are now
     propagated faster.}
  \label{code:ht}
\end{figure}
\normalsize

@@

\begin{figure}[ht!]
   \begin{center}
      \subfloat[]{\includegraphics[width=5cm]{results/new-heat-transfer-80}}
   \end{center}
   \caption{Experimental results for the HT program. Speedup values for the
      coordinated and uncoordinated version are computed using the run time of
      the uncoordinated version using 1 thread.}
   \label{results:ht}
\end{figure}

>>>>

In the Heat Transfer algorithm~(HT), we have a graph where heat values are
exchanged between nodes. The goal of the program is to exchange heat to a point
where new heat value $\delta = |H_i -
H_{i-1}| \le \epsilon$ for all nodes of the graph. The algorithm works
asynchronously since heat values can be updated by using new partial information
coming from neighbor nodes. This increases parallelism since nodes do not need
to synchronize between iterations.

Fig.~\ref{code:ht} shows the HT logical rules that send new heat values to
neighbor nodes. In the first rule we added \texttt{add-priority} to increase the priority of the neighbor
nodes if the current node has a large $\delta$. The idea is to prioritize the
computation of heat values of nodes (using \texttt{update}) that have a neighbor
that changed significantly. Multiple \texttt{add-priority} facts will
increase the priority of a node so that nodes with multiple large deltas will
have more priority. To further improve scalability, we added the third rule to
ignore to not send small $\delta$ values if the target node is in another
thread. Note that the LM compiler will merge the three rules in
Fig.~\ref{code:ht} into a single one that handles the 3 branches.

\begin{figure}[h!]
\scriptsize\begin{Verbatim}[numbers=left,commandchars=\\\[\]]
new-heat(A, New, Old),
Delta = fabs(New - Old),
Delta > epsilon
   -o {B, W | !edge(A, B, W) |
         new-neighbor-heat(B, A, New),
         update(B), \underline[add-priority(B, Delta)]}.

new-heat(A, New, Old)
fabs(New - Old) <= epsilon
\underline[cpu-id(A, B, C)],
\underline[cpu-id(A, A, C)]
   -o {B, W | !edge(A, B, W) |
         new-neighbor-heat(B, A, New)}.

new-heat(A, New, Old)
fabs(New - Old) <= epsilon,
\underline[cpu-id(A, B, C1)],
\underline[cpu-id(A, A, C2)],
\underline[C1 <> C2]
   -o 1. // nothing is derived
\end{Verbatim}
  \caption{Coordination code for the Heat Transfer program. We updated the first rule
     to increase the priority of neighbor nodes and added the third rule to not
     send small $\delta$ if the neighbor is in another thread. The code logic remains exactly
     the same as before, however bigger changes in heat values are now
     propagated faster.}
  \label{code:ht}
\end{figure}
\normalsize

@@

\begin{figure}[ht!]
   \begin{center}
      \subfloat[]{\includegraphics[width=5cm]{results/new-heat-transfer-80}}
   \end{center}
   \caption{Experimental results for the HT program. The coordinated version
      is, on average, 30\% faster than the regular version although it has a
      slightly worse scalability due to a reduction in work available.}
   \label{results:ht}
\end{figure}

### 2014-11-08 implementation.tex
More updates.

The compiler translates each rule to a procedure and a list of facts that need
to exist to satisfy the rule. This procedure is executed by the virtual
machine whenever we have enough facts to satisfy the rule. The procedure
loops over all possible combinations of the rule, retrieving facts from the
database, performing join operations and then consuming and deriving facts.
Optimizations such as join optimizations (to support efficient rule filtering)
and fact updates are implemented by the compiler. The latter
transforms fact derivations of the same predicate to a simple update operation,
avoiding a deallocation operation followed by an allocation of the new fact.

@@

Coordination directives are compiled in two different ways, depending on whether they
appear in the body or in the head of the rule. Coordination facts in the body
are compiled into special instructions that inspect the state of the virtual
machine. For example, the directive \texttt{node-priority} will inspect the
target node, retrieve the current priority and assign the priority to a
register. Coordination facts in the head of the rule are also implemented as
special instructions of the virtual machine, but they perform some action,
instead of returning something. While these are still facts from the point of
view of the language, they are deleted immediatelly in order to apply the
operation, instead of being derived like any other regular fact. This is more
efficient since we do not need to create a new fact.

@@

The virtual machine is implemented in C++11 and uses the threading system from
the standard library to implement multithreading. Each thread is responsible
for executing a subset of nodes. Nodes are either inactive, with no new facts,
or active, where they will be placed in the corresponding thread's queue.
Threads do useful work by processing the nodes in their queues. Whenever a
thread becomes idle, it attempts to steal nodes from a random thread into its
own queues. If sucessful, the thread becomes idle and waits for program
termination while periodically attempting to steal nodes.

@@

A node is represented as a collection of facts (per predicate) and an indexing structure that
keeps track of the available predicates and potential candidate rules. We need
a separate indexing structure per node since rules run locally.

@@

\begin{itemize}
\item \emph{Tries} are used exclusively to store 
  persistent facts. Tries are trees where facts are indexed by common
  prefix arguments.
\item \emph{Doubly Linked Lists} are used to store 
  linear facts. We use a double linked list because it is a very
   efficient way to add and remove facts.
\item \emph{Hash Tables} are used to improve lookup when 
  linked lists are too long and when we need to do search filtered by
  a fixed argument. The virtual machine decides which arguments are
  best to be indexed and then uses a hash table
  indexed by the appropriate argument.
\end{itemize}

@@

\begin{description}
   \item[working]: the node is being executed.
   \item[idle]: the node is not active, has no new facts and is not in any
   queue for processing.
   \item[queue]: the node is active with new facts and is waiting in some queue
   to be processed.
   \item[stealing]: the node has just been stolen is about to move to another
   thread.
   \item[coordinating]: the node is moving from one queue to another (i.e.
         changing priority).
\end{description}

\begin{figure}[h!]
   \centering
   \includegraphics[width=0.4\textwidth]{node_states.pdf}
   \caption{The node state machine. During the lifetime of a program, each node
   goes through different states as specified by the state machine.}
   \label{fig:node_states}
\end{figure}

Each node is protected by a main spinlock that allows threads to change node
attributes: incoming facts, owner thread, node state and locality information.
There is also an internal spinlock that is locked whenever the node enters into
the \textbf{working} state. If a node sends facts to node is placed in
another thread, the current thread first attempts to lock the internal lock in order to update the
indexing structures of the target node, otherwise it adds the facts to the list of
incoming facts that are later processed by the owner thread.

@@

Threads pop active nodes from their queues and execute new candidate rules. 
In order to allow node priorities, we use 2 queues per thread: a doubly linked
list known as the \emph{standard
queue} and a min/max heap known as the \emph{priority queue}. The regular queue
contains nodes without priorities and
implements the operations \texttt{push\_tail(node)} (push to the tail),
\texttt{pop\_head()} (remove node from the head),
\texttt{remove(node)} (remove an arbitrary node) and \texttt{remove\_half()}
(removes first half). The priority queue contains nodes
with priorities and is implemented as a binary heap array. It supports the
operations \texttt{push(node,priority)} (add new node to the heap),
           \texttt{pop\_min()} (remove the best node),
\texttt{remove(node)} (remove an arbitrary node), \texttt{pop\_half\_min()}
(remove the best half) and \texttt{move(node,new\_priority)} (update priority).
Operations such as \texttt{remove\_half()} are supported in order to support
node stealing, while operations \texttt{remove(node)} or
\texttt{move(node,new\_priority)} allows threads to change the priority of
nodes.

@@

Threads synchronize with each other using mutual exclusion. We use a spinlock in
each queue to protect queue operations and one spinlock to protect the thread's
state. Given threads $T_1$ and $T_2$, we enumerate the most important
synchronization intensive places in the virtual machine:

\begin{description}
   \item[New facts:] When a node executes on $T_1$ and derives facts
   to a node in $T_2$, $T_1$ first buffers the facts 
   and then sends them to the target node. Here, it checks if the
   node is currently \textbf{idle} and then synchronizes with $T_2$ to add the
   node to the $T_2$'s queue.
   \item[Thread activation:] If $T_2$ is inactive when adding facts to a node in
   $T_2$, $T_1$ also synchronizes with $T_2$ to change $T_2$'s state to \emph{active}.
   \item[Node stealing:] $T_1$ synchronizes with $T_2$ when it attempts to steal
   nodes from $T_2$ by removing half of the nodes from $T_2$'s queues.
   \item[Coordination:] If $T_1$ needs to perform coordination operations
   to a node in $T_2$, it may need to sychronize with $T_2$ during priority
   updates in order to move the node in $T_2$'s queues.
\end{description}

\subsubsection{Termination}

There is a global atomic counter, a global boolean flag and one boolean flag
(active/idle) per thread for detecting termination. Once a thread goes idle,
it decrements the global counter and changes its flag to idle. If the counter
goes to zero, the global flag is set to idle. Since every thread will be
busy-waiting and checking the global flag, they will detect the change and exit
the program.

\subsubsection{Node collection}

Whenever a node has an empty database and no references from other nodes, it is
deleted from the graph. The virtual machine detects such cases by keeping a
count of fact references to every node. Once the count drops to zero and the
database is empty, the node is immediatelly deleted in order to reduce memory
usage.

@@

Fig.~\ref{results:comparison1} shows the comparison between the original virtual
machine and the new virtual machine with coordination mechanisms. All the
programs benchmarked do not use any kind of coordination. As expected, there
is no degration of performance by adding coordination mechanisms.

\begin{figure}[h!]
   \begin{center}
      \subfloat[]{\includegraphics[width=9cm]{results/comparison1.png}}
   \end{center}
   \caption{Measuring the overhead of coordination mechanisms. The \textbf{Old}
   bars represent the performance of the original virtual machine, while
   \textbf{New} is the relative performance of the new version that includes
   coordination mechanisms.}
   \label{results:comparison1}
\end{figure}

>>>>

The compiler translates each rule to a procedure and a list of facts that need
to exist to satisfy the rule. This procedure is executed by the virtual
machine whenever all the facts mentioned in the body exist.
The procedure loops over all possible combinations of the rule, retrieving facts from the
database, performing join operations and then consuming and deriving facts.
Optimizations such as join optimizations (to support efficient rule filtering)
and fact updates are implemented by the compiler.

@@

Coordination directives are compiled in two different ways, depending on whether they
appear in the body or in the head of the rule. Coordination facts in the body
are compiled into special instructions that inspect the state of the virtual
machine. For example, the directive \texttt{node-priority} will inspect the
target node, retrieve the current priority and assign the priority to a
register. Coordination facts in the head of the rule are also implemented as
special instructions of the virtual machine, but they perform some action,
instead of being added to the database as facts.
Semantically, action facts are like any other. However, since they are
immediatelly used by the machine, there is no need to store them in the
database, avoiding unnecessary allocations and deallocations.

@@

The virtual machine is implemented in C++11 and uses the threading system from
the standard library to implement multithreading. Each thread is responsible for
executing a subset of \emph{active nodes}. A node is active if it has unexamined
facts. After a node is processed, it becomes inactive until a new fact is
derived for it. Inactive nodes have no work to do. When a new fact is derived for an
inactive node, the node is made active and placed on the appropriate queue of
the appropriate thread.  Threads do useful work by processing active nodes from their queues. Whenever a
thread becomes idle, it attempts to steal nodes from a random thread.
If unsucessful, the thread becomes idle and waits for program
termination while periodically attempting to steal nodes.

@@

A node is represented as a collection of facts (per predicate) and an indexing structure that
keeps track of the available facts and potential candidate rules. We need
a separate indexing structure per node since rules run local to each node.

@@

\begin{itemize}
\item \emph{Tries} are used exclusively to store persistent facts.
\item \emph{Doubly Linked Lists} are used to store 
  linear facts. We use a double linked list because we need to efficiently add
  and remove facts.
\item \emph{Hash Tables} are used to improve lookup when 
  linked lists are too long and when we need to do searches filtered by
  a fixed argument. The virtual machine decides which arguments are
  best to be indexed and then uses a hash table
  indexed by the appropriate argument.
\end{itemize}

@@

\begin{description}
   \item[working]: the node is executing.
   \item[inactive]: the node is inactive, i.e., it has no new facts and is not in any
   queue for processing.
   \item[queue]: the node is active with new facts and is waiting in some queue
   to be processed.
   \item[stealing]: the node has just been stolen and is in the process of being
   moved to another thread.
   \item[coordinating]: the node is moving from one queue to another queue.
\end{description}

\begin{figure}[h!]
   \centering
   \includegraphics[width=0.4\textwidth]{node_states.pdf}
   \caption{The node state machine as represented by the state variable. During
      the lifetime of a program, each node goes through different states as
      specified by the state machine.}
   \label{fig:node_states}
\end{figure}

Each node is protected by a main spinlock that allows threads to change node
attributes: incoming facts, owner thread, state variable and locality information.
A database spinlock is locked whenever the node enters into
the \textbf{working} state. If a node sends facts to another node placed in
another thread, the current thread first attempts to lock the database lock in order to update the
indexing structures of the target node, otherwise it adds the facts to the list of
incoming facts that are later processed by the owner thread.

@@

Threads pop active nodes from their queues and execute new candidate rules. 
In order to reduce inter-thread communication, node priorities are implemented
at the thread level. This means that, when a thread picks the highest
priority node, it is actually the highest priority from the set of nodes owned
by the thread and not the highest priority node in the whole program.
To allow node priorities, we use 2 two pairs of queues per thread: a pair of doubly linked
lists known as the \emph{standard queue} and a pair of min/max heap known as the \emph{priority queue}.

The regular queue contains nodes without priorities and
supports push into tail, remove node from the head, remove arbitrary node, and
remove first half of nodes.
The priority queue contains nodes with priorities and is implemented as a binary
heap array. It supports the following operations: push into the heap, remove the min node,
remove an arbitrary node, remove half of the nodes (horizontal split), and
priority update.
Operations for removing half of the queue are implemented in order to support
node stealing, while operations to remove arbitrary nodes or update the
priority of nodes allows threads to change the priority of nodes.

@@

Threads synchronize with each other using mutual exclusion. We use a spinlock in
each queue to protect queue operations.
Given threads $T_1$ and $T_2$, we enumerate the most important
synchronization intensive places in the virtual machine:

\begin{description}
   \item[New facts:] When a node executes in $T_1$ and derives facts
   to a node in $T_2$, $T_1$ first buffers the facts 
   and then sends them to the target node. Here, it checks if the
   node is currently \textbf{idle} and then synchronizes with $T_2$ to add the
   node to the $T_2$'s queue.
   \item[Thread activation:] If $T_2$ is inactive when adding facts to a node in
   $T_2$, $T_1$ also synchronizes with $T_2$ to change $T_2$'s state to \emph{active}.
   \item[Node stealing:] $T_1$ synchronizes with $T_2$ when it attempts to steal
   nodes from $T_2$ by removing half of the nodes from one of $T_2$'s queues.
   \item[Coordination:] If $T_1$ needs to perform coordination operations
   to a node in $T_2$, it may need to sychronize with $T_2$ during priority
   updates in order to move the node in $T_2$'s queues.
\end{description}

\subsubsection{Garbage collection}

Whenever a node has an empty database and no references from other nodes, it is
deleted from the graph. The virtual machine uses reference counting to detect
such cases.

@@

Fig.~\ref{results:comparison1} shows the comparison between the original virtual
machine and the new virtual machine with coordination mechanisms. All the
programs benchmarked do not use any kind of coordination. As hoped for, there
is no degration of performance by adding coordination mechanisms. Since we also
performed general performance improvements, programs such as PageRank and Greedy
Graph Coloring programs now perform better than before.

\begin{figure}[h!]
   \begin{center}
      \subfloat[]{\includegraphics[width=8cm]{results/comparison1.png}}
   \end{center}
   \caption{Measuring the overhead of coordination mechanisms. The \textbf{Old}
   bars represent the performance of the original virtual machine, while
   \textbf{New} is the relative performance of the new version that includes
   coordination mechanisms.}
   \label{results:comparison1}
\end{figure}

### 2014-11-08 intro.tex
More updates.

Writing parallel programs in sequential languages is hard because manipulating
shared state using multiple threads may result in data race
conditions. Such issues are handled with low level constructs such as locks and
condition variables, requiring a fair amount of effort to get right.
Declarative programming has been hailed as a solution to this issue, since the
problem of implementing the details of parallelism is moved away from the programmer to the
compiler and runtime of the language. The programmer writes code
without having to deal with parallel programming constructs and the compiler
automatically parallelizes the program in order to take advantage of
multiple threads of execution.
The programming paradigm has been adopted with huge success in domain specific
languages such as SQL and MapReduce~\cite{Dean:2008:MSD:1327452.1327492}.
Although general declarative languages have yet to be as successful, the
future looks promising for this particular approach.

The problem with declarative programming is that it leaves little to no programmer control
over how execution is scheduled or how data is laid out, making it hard to achieve
necessary efficiency. This introduces some
performance issues because even if the runtime system is able to reasonably
parallelize the program using a general algorithm, there is a lack of specific
information about the program that a compiler does not easily understand. In
turn, such information would make execution better in terms of run time time,
memory usage and scalability.

In this paper, we extend Linear Meld (LM), a declarative language by
\cite{cruz-iclp14,cruz-ppdp14} with coordination facts that give programmer
control over scheduling and data placement. LM is a linear logic programming
language designed for programs that operate on graphs.  The use of linear
logic~\cite{girard-87} allows support for structured manipulation of mutable
state. In LM, computation is divided so that each node of the graph computes
independently and is allowed to communicate with other nodes.
Both computation and communication happen through derivation of logical rules.

Given that LM uses mutable state, the order in which nodes are executed may be
crucial in terms of performance. We extended the language with coordination
facts that can be used as a regular fact in the program logic to improve program
execution.  The first kind of coordination facts are called \emph{sensing facts}
and are used to sense information about the system the program is running on,
e.g., scheduling and node placement on threads. The second kind of coordination facts are
\emph{action facts} that when deleted in rules are used to apply a scheduling
operation during execution. Coordination facts allow the programmer to write
logical rules that depend on the current state of the program and then prioritize node
computation or place nodes in different threads.

To the best of our knowledge, this is the first time that a declarative language allows 
control over execution without a significant change in the language itself and
without using non-declarative operations. This is specially important since
programs can be proven to be correct even in the presence of coordination facts.
In order to prove our thesis, we first give a brief overview of the
base language, including a small example. Next, in
Section~\ref{sec:coordination} we introduce the available coordination facts,
followed by Section~\ref{sec:implementation}, where we describe the
implementation of the runtime required to implement the desired coordination
mechanisms. Afterwards, in Section~\ref{sec:applications} we present several applications where we add
programmer control to declarative programs in order to make them perform better.

>>>>

Writing parallel programs in sequential languages is hard because manipulating
shared state using multiple threads may result in data race
conditions. Such issues are handled with low level constructs such as locks and
condition variables, requiring a fair amount of effort to get right.
Declarative programming has been hailed as a solution to this issue, since the
problem of implementing the details of parallelism is moved from the programmer to the
compiler and runtime. The programmer writes code
without having to deal with parallel programming constructs and the compiler
automatically parallelizes the program in order to take advantage of
multiple threads of execution.
The programming paradigm has been adopted with huge success in domain specific
languages such as SQL and MapReduce~\cite{Dean:2008:MSD:1327452.1327492}.
Although general declarative languages have yet to be as successful, the
future looks promising for this particular approach.

The problem with declarative programming is that it leaves little to no programmer control
over how execution is scheduled or how data is laid out, making it hard to achieve
necessary efficiency. This introduces 
performance issues because even if the runtime system is able to reasonably
parallelize the program using a general algorithm, there is a lack of specific
information about the program that a compiler cannot easily deduce. Such
information could make execution better in terms of run time time,
memory usage, and scalability.

In this paper, we introduce Coordinated Linear Meld (CLM), a declarative
language that extends the Linear Meld~(LM) language~\cite{cruz-iclp14,cruz-ppdp14} with coordination facts that give programmer
control over scheduling and data placement. LM is a linear logic programming
language designed for programs that operate on graphs.  The use of linear
logic~\cite{girard-87} supports structured manipulation of mutable
state. In LM, computation is divided so that each node of the graph computes
independently and is allowed to \scare{communicate} with other nodes.
Both computation and communication happen through the derivation of logical
rules (which make up the program).

The CLM language features coordination
primitives that can be used in the same way as any other primitive,
in other words, they are specificied with the same syntax and semantics as the
rest of the programming language. These coordination primitives can be used to
improve program execution based on the state of the program and the underlying
machine. The first kind of coordination facts are called \emph{sensing facts}
and are used to sense information about the system the program is running on,
e.g., scheduling and node placement on threads. The second kind of coordination facts are
\emph{action facts} that when deleted in rules are used to apply a scheduling
operation during execution. Coordination facts allow the programmer to write
logical rules that depend on the current state of the program and then prioritize node
computation or place nodes in different threads.

To the best of our knowledge, this is the first time that a declarative language allows 
control over execution while staying declarative and without resorting to
meta-language constructs. This is crucial to our goal of being able to prove
programs correct since
proofs can be constructed even in the presence of coordination facts.
After briefly discussing related work, we present an overview of the base
language, with an example. In Section~\ref{sec:coordination}, we introduce the
current set of coordination facts,
followed by a description of the changes required to implement the desired
coordination mechanisms. In
Section~\ref{sec:applications} we present several applications and show
coordination can improve programs without destroying clarity or provability.
